---
title: "Machine Learning Peer-graded Assignment"
author: "A Steenekamp"
date: "10 November 2016"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Executive Summary 
## Background
By using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify *how much* of a particular activity they do, but they rarely quantify *how well* they do it. In this project, 6 participants produce data from accelerometers on the belt, forearm, arm, and dumbell.They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

*Note: Classe A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes.*

Read more on HAR and the Weight-lifting dataset at http://groupware.les.inf.puc-rio.br/har#ixzz4PagQx474

## Objective
The goal of this project is to predict the manner in which the participants did the exercise. This is the "classe" variable in the training set.A report describing how the model was built, how cross validation was used, what the expected out of sample error is, and why certain choices were made. The prediction model will be used to predict 20 different test cases.

# Setup environments
## Load libraries

```{r load libraries}
suppressMessages(library(randomForest))
suppressMessages(library(knitr))
suppressMessages(library(ggplot2))
suppressMessages(library(datasets))
suppressMessages(library(corrplot))
suppressMessages(library(caret))
suppressMessages(library(rpart.plot))
suppressMessages(library(rattle))
suppressMessages(library(rpart))
suppressMessages(library(gbm))
```
# Getting and Cleaning the data
## Read the data
```{r read files}
cache=TRUE
Work_Dir <- "C:/Users/Ada/Documents/Coursera/Mod8MachineLearning"
setwd(Work_Dir)
if (!file.exists("pml-training.csv")) {
    download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
                  destfile = paste(Work_Dir, "pml-training.csv", sep="/"))
}
if (!file.exists("pml-testing.csv")) {
    download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
                  destfile = paste(Work_Dir, "pml-testing.csv", sep="/"))
}
traindata <- data.frame(read.csv("pml-training.csv",na.strings=c("", "NA", "NULL")))
testdata <- data.frame(read.csv("pml-testing.csv",na.strings=c("", "NA", "NULL")))
```
## Explore the original data
```{r investigate data}
dim(traindata)
dim(testdata)
```
There is a very large number of variables and we need to determine which are the most important ones and try and reduce the dataset so that only these are left.

## Create a new dataset with only the most important variables
### 1) Manual inspection
By manually looking at the data it is clear that we can remove the first 7 columns.They are identifying data and will not contribute to the correctness of the exercise
```{r remove first 7 columns}
traindata<-traindata[-c(1:7)]
dim(traindata)
```
### 2) Remove columns with too many NA's
```{r remove NAs}
set.seed(12345)
trainNA <- traindata[ , colSums(is.na(traindata)) == 0]
dim(trainNA)
```
### 3) Determine highly correlated values
```{r determine correlation}
# first we need to remove the "classe" variable, since the cor function can only be performed on numeric values
trainnum <- trainNA[,-c(53)]
cortrain<-cor(trainnum)
```
### 4) Remove highly correlated values
```{r remove cors}
removecor = findCorrelation(cortrain, cutoff = .90, verbose = FALSE)
trainrem = trainnum[,-removecor]
dim(trainrem)
cortrain<-cor(trainrem)
corrplot(cortrain,method="circle",order = "FPC", type = "lower", tl.cex = 0.6)
```


### 5) Add classe variable back into the dataset
```{r add classe variable}
trainfin=trainNA[,-removecor]
dim(trainfin)
```
### 6) Create final training and testing datasets
```{r create training and testing datasets}
inTrain = createDataPartition(y=trainfin$classe,p = 3/4,list=FALSE)
training=trainfin[inTrain,]
testing=trainfin[-inTrain,]
dim(training)
dim(testing)
```
# Apply machine learning algorithms
## 1) Decision Tree (method="rpart")
```{r first method}
set.seed(12345)
fit1<-train(classe~.,method="rpart",data=training)
fancyRpartPlot(fit1$finalModel)
predrpart<-predict(fit1,testing)
confusionMatrix(predrpart,testing$classe)$overall[1]
```
## 2) Random Forest (method="rf")
```{r second method}
set.seed(12345)
fit2<-randomForest(classe~.,data=training,ntree=100, importance=TRUE)
predrf<-predict(fit2,testing)
confusionMatrix(predrf,testing$classe)$overall[1]
```
## 3) Generalized Boosted Regression Models (method = "gbm")
```{r third method}
set.seed(12345)
controlgbm <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
fit3<-train(classe~.,method="gbm",data=training,verbose=FALSE,trControl = controlgbm)
predgbm<-predict(fit3,testing)
confusionMatrix(predgbm,testing$classe)$overall[1]
```
# Predict the test cases
The accuracy rate for the three methods are:
1) Decision Tree 54%, 51% expected out of error rate
2) Random Forest 99%, 1% expected out of error rate
3) Boosting 96%, 4% expected out of error rate
The most accurate method is **Random Forest** and this will be used to predict the test cases
```{r predict the test cases}
predtest <- predict(fit2, newdata=testdata)
predtest
```
